---
title: "Plasticity and adherence"
author: "Jess Grembi"
date: "11/9/2018"
output: 
  BiocStyle::html_document:
      number_sections: no
      toc: yes
      toc_float: yes
---


# Setup

```{r setup}
#rm(list = ls())

.packages <- c("gridExtra", "ggplot2", "viridis",  "RColorBrewer", 
               "knitr",  "phyloseq", "tidyverse", "ggsignif", "ggbeeswarm")
sapply(.packages, require, character.only = TRUE)

options(stringsAsFactors = FALSE)
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE, 
  fig.path = "../figs/plasticity/", dev= c('png', 'pdf'),
  fig.align = 'center', fig.wide = TRUE, 
  fig.width = 10, fig.height = 8
)


theme_set(theme_bw())
theme_update(
  text = element_text(size = 20),
  strip.background = element_rect(fill="grey97")
)
```



## Load data

```{r}
(ps0 <- readRDS("../data/phyloseq/diet_participants_with_adherence.rds"))

# calculate adherence, and Subj_Group variables
newSam_data <- sample_data(ps0) %>%
  data.frame() %>%
  mutate_at(vars(starts_with("adherence")), as.numeric) %>%
  mutate(adherence.raw = rowMeans(select(., adherence_3_months, adherence_6_months, adherence_12_months), na.rm = T),
         adherence.pct = 100-adherence.raw, 
         adherence.changeFromBL = adherence_baseline - adherence.raw,
         Subj_Group = factor(paste0(SubjectID, "_", Timepoint)), 
         rowname = MeasID) %>%
  column_to_rownames("rowname")


sample_data(ps0) <- newSam_data

```


## Filter out samples with insufficient reads or no weight-loss data at 12 months

```{r}
minSampleDepth <- 10000


(ps0 <- subset_samples(ps0, sample_sums(ps0) > minSampleDepth))
(ps0 <- subset_samples(ps0, !(SubjectID == "1139" & Timepoint == "10 Week")))
(ps0 <- subset_samples(ps0, !is.na(Cat.12Mo)))
ps0 <- subset_taxa(ps0, taxa_sums(ps0) > 0)

```

## Twice-sequenced samples

Find biological samples which were sequenced twice:

```{r}
(sample_data(ps0) %>% 
   group_by(SampleID) %>% 
   dplyr::summarise(
     count = n(), 
     meas = paste0(MeasID, collapse = "; ")
     ) %>%
   filter(count > 1))

```


Combine repeated samples by computing the mean 

```{r}
collapse_samples <- function(physeq, sample_id = "SampleID", fun = mean){
  sampledata_collapsed <- sample_data(physeq) %>%
    dplyr::select(-Lane, -MeasID, -OrigSampleDepth) %>%
    distinct() 
  rownames(sampledata_collapsed) <- sampledata_collapsed[[sample_id]]
  seqtab <- data.frame(as(otu_table(physeq), "matrix"))
  if(!taxa_are_rows(physeq)) seqtab <- t(seqtab)
  
  sample_ids <- physeq@sam_data[[sample_id]]
  seqtab_collapsed <- sapply(unique(sample_ids), function(id){ 
    if(sum(sample_ids == id) == 1) return(seqtab[, sample_ids == id])
    apply(seqtab[, sample_ids == id], 1, fun)
  })
  rownames(seqtab_collapsed) <- rownames(seqtab)
  physeq_collapsed <- phyloseq(
    otu_table(seqtab_collapsed, taxa_are_rows = TRUE),
    sample_data(sampledata_collapsed), 
    tax_table(physeq), phy_tree(physeq))
  return(physeq_collapsed)
}
```



```{r}
ps0 <- collapse_samples(ps0)
ps0
sample_data(ps0)$Cat.12Mo <- 
  factor(sample_data(ps0)$Cat.12Mo, levels = c("US", "MS", "VS"))
```



# Plasticity/Variability compared to weight loss

```{r}
### Function to calculate beta diversity
## Modified from Lan's original code to allow all comparisons to be made at once.  Then we can filter later
beta_diversity <- function(physeq, timediff_thresh = 5,
                           sample_info_cols = c("Diet", "Cat.12Mo", "SampleDate", "Percent.12Mo", "Timepoint", "adherence.pct", "adherence.changeFromBL", "Gender"),
                           asinh = TRUE, metric = "bray") {
  physeq <- phyloseq::prune_taxa(taxa_sums(physeq) > 0, physeq)
  X <- as(otu_table(physeq), "matrix")
  if(taxa_are_rows(physeq)) X <- t(X)
  if(asinh) {
    X <- asinh(X)
    physeq <- phyloseq::phyloseq(
      otu_table(X, taxa_are_rows = FALSE),
      sample_data(physeq), phy_tree(physeq))
  }
  smp_df <- sample_data(physeq) %>%
    data.frame() %>%
    dplyr::select("SampleID", "SubjectID", sample_info_cols) %>%
    mutate_at(
      .funs = as.character, 
      .vars = c("SampleID", "SubjectID")
    )
  if (metric == "jaccard") {
    D <- dist(floor(X), method = "binary")
  } else if (metric == "jsd") {
    D <- phyloseq::distance(physeq, method = "jsd")
  } else if (metric %in% c("unifrac", "wunifrac")) {
    D <- UniFrac(physeq, weighted = ifelse(metric == "wunifrac", TRUE, FALSE))
  } else {
    D <- vegan::vegdist(X, method = metric)
  }
  df_dist <- reshape2::melt(
    as.matrix(D), 
    varnames = c("S1", "S2"),
    value.name = "dist") %>%
    mutate(
      S1 = factor(S1, levels = smp_df$SampleID, ordered = TRUE), 
      S2 = factor(S2, levels = smp_df$SampleID, ordered = TRUE)) %>%
    filter(S1 < S2) %>%
    mutate(
      S1 = as.character(S1),
      S2 = as.character(S2)) %>%
    left_join(smp_df, by = c("S1" = "SampleID")) %>%
    left_join(smp_df, by = c("S2" = "SampleID"),suffix = c("_1", "_2")) %>%
    filter(SubjectID_1 == SubjectID_2) %>%
    mutate(SubjectID = SubjectID_1) %>%
    dplyr::select(-SubjectID_1, -SubjectID_2)
    
    if("SampleDate" %in% colnames(smp_df)) {
      df_dist <- df_dist %>%
        mutate(
          SampleDate_1 = as.Date(SampleDate_1),
          SampleDate_2 = as.Date(SampleDate_2),
          TimeDiff = as.numeric(SampleDate_2 - SampleDate_1)
        ) %>%
        filter(abs(TimeDiff) <= timediff_thresh) %>%
        # group_by(SubjectID) %>%
        # top_n(2, wt = -TimeDiff) %>%
        as.data.frame()
    }
  
    cols <- setdiff(sample_info_cols, c("SampleDate", "Timepoint"))

    df_dist <- df_dist %>%
      dplyr::rename_at(
        .funs = function(x) {gsub("_1", "", x)},
        .vars = paste0(cols, "_1"))
    
    df_dist <- df_dist %>%
      dplyr::select(setdiff(colnames(df_dist), paste0(cols, "_2")))
  return(df_dist)
}

```


## Calculate beta diversity between all samples

```{r, eval = FALSE}
dist_lst <- c("jaccard", "unifrac", "jsd", "bray", "wunifrac")
beta_diver_lst <- lapply(dist_lst, function(dist) {
  beta_diversity(
    physeq  = ps0,
    metric = dist,
    timediff_thresh = Inf,
    asinh = TRUE)})
names(beta_diver_lst) <- dist_lst
saveRDS(beta_diver_lst, file = "../results/beta_diver.rds")
```

```{r}
dist_lst <- c("jaccard", "unifrac", "jsd", "bray", "wunifrac")
beta_diver_lst <- readRDS("../results/beta_diver.rds")
```


```{r}
beta_diver_df <- plyr::ldply(
  beta_diver_lst, function(df) df, .id = "metric") %>%
  mutate(metric = factor(
    metric, levels = dist_lst, 
    labels = c("Jaccard", "UniFrac", "Jensen-Shannon", "Bray-Curtis",
               "Weighted UniFrac")),
    Cat.12Mo = factor(Cat.12Mo, levels = c("US", "MS", "VS"))) %>%
  left_join(ps0@sam_data %>% 
          dplyr::select(SubjectID, Cohort) %>% 
          mutate(
            SubjectID = as.character(SubjectID)) %>% 
          filter(!duplicated(SubjectID)))
```


### Number of days between samples
```{r}

##Stats on # days between samples at same timepoint
(TimeDiff.stats <- beta_diver_df %>% 
  filter(Timepoint_1 == Timepoint_2) %>%
  group_by(Timepoint_1, Cohort) %>%
  summarise(medianTD = median(TimeDiff), pct.25 = quantile(TimeDiff, 0.25), 
            pct.75 = quantile(TimeDiff, 0.75)))

##Stats on # days between samples at BL and 10-weeks 
(TimeDiff.stats <- beta_diver_df %>% 
    filter(Cohort == 2, metric == "Jaccard") %>%
    filter(Timepoint_1 != Timepoint_2) %>%
    group_by(Diet) %>%
    summarise(medianTD = median(abs(TimeDiff)), 
              pct.25 = quantile(abs(TimeDiff), 0.25), 
              pct.75 = quantile(abs(TimeDiff), 0.75), 
              meanTD = mean(abs(TimeDiff)), 
              sdTD = sd(abs(TimeDiff))))
```

Cohort 3 samples are not collected on consecutive days so assessing daily plsticity is not feasible. 
In addition, no samples are available from 10-weeks so we are unable to assess plasticity between baseline and 10-weeks.

##Daily baseline (pre-diet) variability
Keeping only consecutive samples that are within 5 days of each other (a few subjects did not follow instructions and samples are farther apart!)

```{r}
beta_diver_df_bl <- beta_diver_df %>%
  filter(Cohort == 2,
         Timepoint_1 == Timepoint_2,
         Timepoint_1 == "Baseline",
         abs(TimeDiff) <= 5) %>%
  group_by(SubjectID, metric) %>%
  top_n(2, wt = -TimeDiff) 


beta_diver_df_bl_subj <- beta_diver_df_bl %>%
  data.frame() %>%
  group_by(metric, SubjectID, Diet, Gender) %>%
  dplyr::summarise(
    dist = mean(dist),
    Cat.12Mo = unique(Cat.12Mo), Percent.12Mo = unique(Percent.12Mo))
```



```{r}
table(beta_diver_df_bl_subj$Diet, beta_diver_df_bl_subj$metric)
```


```{r short-term-var-diet-bl-seqs_SI, fig.width=6, fig.height=12}
diet_cols <- c(brewer.pal(9, "Set1")[2], brewer.pal(9, "Set1")[5])
ggplot(beta_diver_df_bl_subj) +
  geom_boxplot(
    aes(y = dist, x = Cat.12Mo, color = Diet)
  ) +
  geom_jitter(
    data = beta_diver_df_bl,
    aes(y = dist, x = Cat.12Mo),
    width = 0.15,
    height = 0,
    color = "grey55", size = 1
  ) +
  geom_jitter(
    aes(y = dist, x = Cat.12Mo, color = Diet),
    width = 0.05, height = 0,
    shape = 17, size = 3
  ) +
  stat_signif(
    data = beta_diver_df_bl_subj, 
    aes(y = dist, x = Cat.12Mo),
    stat = "signif",
    comparisons = list(c("VS", "US")),
    test = "wilcox.test", 
   # test.args = list(alternative = "greater"),
    manual = FALSE,
    margin_top = -.12 , 
    map_signif_level = F, 
    vjust = -0.1
  ) +
  facet_grid(metric ~ Diet, scales = "free") +
  scale_color_manual(values =  diet_cols) +
  labs(y = "Pre-diet daily microbiota plasticity", 
       x = "12-month weight-loss success category") +
  theme(legend.position="none")
```


```{r}
wilcoxon_test <- lapply(unique(beta_diver_df_bl_subj$metric), function(m) {
  df <- beta_diver_df_bl_subj %>%
    filter(metric == m)
  
  wtest_lst <- lapply(unique(df$Diet), function(diet) {
    diet_df <- df %>% filter(Diet == diet)
    max_y = max(df$dist)
    diet_wtest <- wilcox.test(
      x = diet_df %>% filter(Cat.12Mo == "VS") %>% .[["dist"]], 
      y = diet_df %>% filter(Cat.12Mo == "US") %>% .[["dist"]], 
      alternative = "two.sided") 
    diet_wtest <- broom::tidy(diet_wtest) %>%
      mutate(max_y = max_y)
  })
  names(wtest_lst) <- unique(df$Diet)
  wtest_df <- plyr::ldply(wtest_lst, function(df) df, .id = "Diet")
})
names(wilcoxon_test) <- unique(beta_diver_df_bl_subj$metric)
wilcoxon_test <- plyr::ldply(wilcoxon_test, function(df) df, .id = "metric") %>%
  mutate(xmin = "US", xmax = "VS", 
         annotation = paste0("p=", signif(p.value,2)), y_position = max_y + 0.05)
wilcoxon_test %>% arrange(Diet, metric)
```

```{r}
spearman_corr <- beta_diver_df_bl_subj %>%
  mutate(Percent.12Mo = as.numeric(Percent.12Mo)) %>%
  group_by(Diet, metric) %>%
  mutate(
    rank_dist = rank(dist),
    rank_weight_loss = rank(Percent.12Mo))

spearman_corr %>%
  summarise(spearman = cor(rank_weight_loss, rank_dist, method = "spearman"))

```




## Daily plasticity at 10-weeks
Keeping only consecutive samples that are within 5 days of each other (a few subjects did not follow instructions and samples are farther apart!)

```{r}
beta_diver_df_10wk <- beta_diver_df %>%
  filter(Cohort == 2,
         Timepoint_1 == Timepoint_2,
         Timepoint_1 == "10 Week",
         abs(TimeDiff) <= 5) %>%
  group_by(SubjectID, metric) %>%
  top_n(2, wt = -TimeDiff) 


beta_diver_df_10wk_subj <- beta_diver_df_10wk %>%
  data.frame() %>%
  group_by(metric, SubjectID, Diet) %>%
  dplyr::summarise(
    dist = mean(dist),
    Cat.12Mo = unique(Cat.12Mo), Percent.12Mo = unique(Percent.12Mo))
```


```{r}
table(beta_diver_df_10wk_subj$Diet, beta_diver_df_10wk_subj$metric)
```


```{r short-term-var-10week-diet-seqs, fig.width=6, fig.height=12}
ggplot(beta_diver_df_10wk_subj) +
  geom_boxplot(
    aes(y = dist, x = Cat.12Mo, color = Diet)
  ) +
  geom_jitter(
    data = beta_diver_df_10wk,
    width = 0.15, height = 0,
    aes(y = dist, x = Cat.12Mo),
    color = "grey55", size = 1
  ) +
  geom_jitter(
    aes(y = dist, x = Cat.12Mo, color = Diet),
    width = 0.05, height = 0,
    shape = 17, size = 3
  ) +
  stat_signif(
    data = beta_diver_df_10wk_subj, 
    aes(y = dist, x = Cat.12Mo),
    stat = "signif", 
    comparisons = list(c("VS", "US")), 
    test = "wilcox.test", 
   # test.args = list(alternative = "greater"), 
    manual = FALSE, 
    margin_top = -.12 , 
    map_signif_level = FALSE, 
    vjust = -0.1
  ) + 
  facet_grid(metric ~ Diet, scales = "free") +
  scale_color_manual(values =  diet_cols) + 
  labs(y = "10-week daily plasticity",
       x = "12-month weight loss success category") + 
  theme(legend.position="none")

```


## Plasticity from pre-diet to 10 weeks

```{r}
beta_diver_df_longTerm <- beta_diver_df %>%
  filter(Cohort == 2, Timepoint_1 != Timepoint_2) 


beta_diver_df_longTerm_subj <- beta_diver_df_longTerm %>%
  group_by(metric, SubjectID, Diet, Cat.12Mo,Percent.12Mo) %>%
  dplyr::summarise(dist = mean(dist)) 
```


```{r}
table(beta_diver_df_longTerm_subj$Diet, beta_diver_df_longTerm_subj$metric)
```


```{r long-term-var-diet-seqs_pairwise, fig.height=12, fig.width=6}
ggplot(data = beta_diver_df_longTerm_subj) +
  geom_boxplot(
    aes(y = dist, x = Cat.12Mo, color = Diet)
  ) +
  geom_jitter(
    data = beta_diver_df_longTerm,
    width = 0.15, height = 0,
    aes(y = dist, x = Cat.12Mo),
    color = "grey55", size = 1
  ) +
  geom_jitter(
    aes(y = dist, x = Cat.12Mo, color = Diet),
    width = 0.05, height = 0,
    shape = 17, size = 3
  ) +
  stat_signif(
    data = beta_diver_df_longTerm_subj, 
    aes(y = dist, x = Cat.12Mo),
    stat = "signif", 
    comparisons = list(c("VS", "US")), 
    test = "wilcox.test", 
   # test.args = list(alternative = "greater"), 
    manual = FALSE, 
    margin_top = -.12 , 
    map_signif_level = FALSE, 
    vjust = -0.1
  ) + 
  facet_grid( metric ~ Diet, scales = "free") +
  scale_color_manual(values = diet_cols) + 
  labs(y = "Microbiota plasticity between pre-diet and 10-weeks", 
       x = "12-month weight loss success category") + 
  theme(legend.position="none")
  
```



```{r}
wilcoxon_test <- lapply(unique(beta_diver_df_longTerm_subj$metric), function(m) {
  df <- beta_diver_df_longTerm_subj %>%
    filter(metric == m)
  
  wtest_lst <- lapply(unique(df$Diet), function(diet) {
    diet_df <- df %>% filter(Diet == diet)
    diet_wtest <- wilcox.test(
      x = diet_df %>% filter(Cat.12Mo == "VS") %>% .[["dist"]], 
      y = diet_df %>% filter(Cat.12Mo == "US") %>% .[["dist"]], 
      alternative = "two.sided") 
    diet_wtest <- broom::tidy(diet_wtest)
  })
  names(wtest_lst) <- unique(df$Diet)
  wtest_df <- plyr::ldply(wtest_lst, function(df) df, .id = "Diet")
})
names(wilcoxon_test) <- unique(beta_diver_df_longTerm_subj$metric)
wilcoxon_test <- plyr::ldply(wilcoxon_test, function(df) df, .id = "metric")
wilcoxon_test %>% arrange(Diet, metric)
```


```{r}
spearman_corr <- beta_diver_df_longTerm_subj %>%
  mutate(Percent.12Mo = as.numeric(Percent.12Mo)) %>%
  group_by(Diet, metric) %>%
  dplyr::summarise(spearman = cor(Percent.12Mo, dist, method = "spearman"),
                   pearson = cor(Percent.12Mo, dist, method = "pearson"))
spearman_corr
```


```{r}
beta_diver_df_bl <- beta_diver_df %>%
  filter(Cohort == 2,
         Timepoint_1 == Timepoint_2,
         Timepoint_1 == "Baseline",
         abs(TimeDiff) <= 5) %>%
  group_by(SubjectID, metric) %>%
  top_n(2, wt = -TimeDiff) 
```

## Comparison of Microbiota plasticity across all time points


```{r}
beta_diver_df <- beta_diver_df %>%
  filter(
    Cohort == 2,
    !(Timepoint_1 == Timepoint_2 & TimeDiff > 5 )) %>%
  mutate(VarCat = factor(
    ifelse(Timepoint_1 == Timepoint_2,
          ifelse(Timepoint_1 == "Baseline", "BL v. BL", "10wk v. 10wk"), 
          "BL v. 10wk"), 
  levels = c( "BL v. BL", "10wk v. 10wk", "BL v. 10wk"))) 


beta_diver_df <- rbind(
  data.frame(beta_diver_df %>% filter(VarCat != "BL v. 10wk") %>%
     group_by(SubjectID, metric, VarCat) %>%
     top_n(2, wt = -TimeDiff)), # pick only distances from (A,B) and (B, C) and NOT (A,C) 
                               # when evaluating  "BL v. BL", "10wk v. 10wk"
  data.frame(beta_diver_df %>% filter(VarCat == "BL v. 10wk"))
)

 

beta_diver_df_bySubj <- beta_diver_df %>%
  group_by(SubjectID, Gender, Diet, VarCat, metric) %>%
  dplyr::summarise(
    dist = mean(dist), meanTimeDiff = mean(TimeDiff),
    medianTimeDiff = median(TimeDiff),
    Cat.12Mo = unique(Cat.12Mo), 
    Percent.12Mo = unique(Percent.12Mo),
    adherence.pct = unique(adherence.pct),
    adherence.changeFromBL = unique(adherence.changeFromBL))

(beta_diver_Summary <- beta_diver_df_bySubj %>%
  filter(metric == "Bray-Curtis") %>%
  group_by(Diet, VarCat) %>%
  dplyr::summarise(meanTD = mean(meanTimeDiff), 
                   medianTD = median(meanTimeDiff)))

```

```{r variability_acrossAllTimes-facet, fig.height=8, fig.width=12}
ggplot(
  data = beta_diver_df_bySubj %>% 
    filter(metric %in% c("Bray-Curtis")) %>%
    ungroup() %>%
    mutate(
      VarCat = factor(VarCat, levels = c("BL v. 10wk", "10wk v. 10wk", "BL v. BL"))), 
  aes(x = Cat.12Mo, y = dist)
  ) + 
  geom_boxplot(
    aes(color = Cat.12Mo),
    outlier.shape = NA
  ) + 
  geom_point(
    data = beta_diver_df %>% filter(metric %in% c("Bray-Curtis")), 
    aes( x = Cat.12Mo, y = dist, fill = Cat.12Mo), 
    position = position_jitterdodge(jitter.width = 0.5), 
    color = "grey55", 
    size = 1.5, 
    alpha = 0.5, 
    show.legend=FALSE
  ) + 
  geom_point(
    aes(color = Cat.12Mo),
    position = position_jitterdodge(jitter.width = 0.3), 
    shape = 17, 
    size = 3
  ) +
  stat_signif(
    stat = "signif",
    label = "p.adj",
    comparisons = list(c("VS", "US")),
    test = "wilcox.test",
    margin_top = -.12 ,
    vjust = -0.5
  ) +
  facet_grid(VarCat ~ Diet) + 
  scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) + 
  labs(x = " ", y = "Bray-Curtis dissimilarity", color = "") + 
  coord_flip() + 
  guides(color = guide_legend(reverse=T, title = "Weight Loss Success")) +
  theme(text = element_text(size = 20), legend.position = "top")

```

The ones below are probably adjusted p-values


```{r variability_acrossAllTimes, fig.height=6, fig.width=12}
library(ggpubr)
ggplot(
  data = beta_diver_df_bySubj %>% 
    filter(metric == "Bray-Curtis"), 
  aes(x = VarCat, y = dist)
  ) + 
  geom_boxplot(
    aes(color = Cat.12Mo),
    outlier.shape = NA
  ) + 
  geom_point(
    data = beta_diver_df %>% filter(metric == "Bray-Curtis"), 
    aes(x = VarCat, y = dist, lty = Cat.12Mo), 
    position = position_jitterdodge(jitter.width = 0.25), 
    color = "grey70", size = 1.5, alpha = 0.5
  ) + 
  geom_point(
    aes(color = Cat.12Mo),
    position = position_jitterdodge(jitter.width = 0.1), 
    shape = 17, size = 3
  ) +
  stat_compare_means(
    data = beta_diver_df_bySubj %>%
      filter(metric %in% c("Bray-Curtis"), Cat.12Mo != "MS"),
    mapping = aes(x = VarCat, y = dist, group = Cat.12Mo), 
    size = 6,
    label.y.npc = 0.75, 
    label = "p.format", 
    method = "wilcox.test", 
    method.args = list(alternative = "two.sided"),
    tip.length = 0.05,
    inherit.aes = FALSE) +
  # stat_signif(
  #   stat = "signif",
  #   label = "p.adj",
  #   comparisons = list(c("VS", "US")),
  #   test = "wilcox.test",
  #   margin_top = -.12 ,
  #   vjust = -0.5
  # ) +
  facet_grid(~Diet, scales = "free") +
  scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) + 
  labs(x = " ", y = "Bray-Curtis dissimilarity", color = "") + 
  coord_flip() + 
  guides(color = guide_legend(reverse=T, title = "Weight Loss Success")) +
  theme(text = element_text(size = 25), legend.position = "top")
```


No difference in Microbiota plasticity between men and women at any time point (BL v BL, 10wk v 10wk, or BL v 10wk). 

Also looked by diet and differences between men and women by diet were also insignificant at each time point.

```{r, fig.height=6, fig.width=10}

ggplot(beta_diver_df_bySubj, 
       aes(x = Gender, y = dist)
) + 
  geom_boxplot() + 
  stat_signif(stat = "signif",
              comparisons = list(c("Male", "Female")),
              test = "wilcox.test", 
              test.args = list(alternative = "two.sided"),
              manual = FALSE,
              margin_top = -.12 , map_signif_level = FALSE, vjust = -0.1
  ) +
  facet_wrap(VarCat~metric, scales = "free_y", ncol = 5) + 
  labs(x = "", y = "Microbiota plasticity")


```

Is Pre-diet/baseline plasticity (BL v BL) correlated to plasticity over 10 weeks (BL v 10wk)?

```{r, fig.height=6, fig.width=10}

beta_diver_df_bySubj_wide <- beta_diver_df_bySubj %>%
  select(-meanTimeDiff, -medianTimeDiff) %>%
  reshape2::dcast(SubjectID + metric + adherence.pct + adherence.changeFromBL + Diet + Cat.12Mo ~ VarCat, value.var = "dist") %>%
  rename(plas.BL = `BL v. BL`,
         plas.10wk = `10wk v. 10wk`, 
         plas.BL_10wk = `BL v. 10wk`)



ggplot(data = beta_diver_df_bySubj_wide, aes( x = plas.BL, y = plas.BL_10wk)) + 
  geom_point(aes(color = Cat.12Mo), size = 3) + 
  ggpubr::stat_cor(method = "spearman") +
  facet_wrap(Diet~metric, ncol = 5) + 
  labs(x = "BL plasticity", y = "BL-->10wk plasticity", color = "") + 
  stat_smooth(method = "lm") + coord_fixed(ratio = 1)



ggplot(data = beta_diver_df_bySubj_wide, aes( x = plas.BL, y = plas.10wk)) + 
  geom_point(aes(color = Cat.12Mo), size = 3) + 
  ggpubr::stat_cor(method = "spearman") +
  facet_wrap(Diet~metric, ncol = 5) + 
  labs(x = "BL plasticity", y = "10wk plasticity", color = "") + 
  stat_smooth(method = "lm") + coord_fixed(ratio = 1)


ggplot(data = beta_diver_df_bySubj_wide, aes( x = plas.10wk, y = plas.BL_10wk)) + 
  geom_point(aes(color = Cat.12Mo), size = 3) + 
  ggpubr::stat_cor(method = "spearman") +
  facet_wrap(Diet~metric, ncol = 5) + 
  labs(x = "10wk plasticity", y = "BL-->10wk plasticity", color = "") + 
  stat_smooth(method = "lm") + coord_fixed(ratio = 1)
```



### Microbiota plasticity vs alpha-diversity

```{r}
blue_curves_df <- readRDS("../data/alphaDiversity/diet_phylocurves_blue_cohort2.rda")
purp_curves_df <- readRDS("../data/alphaDiversity/diet_phylocurves_purple_cohort2.rda")

alphadiver_cohort2 <- bind_rows(blue_curves_df, purp_curves_df) %>%
  filter(floor(m) == 11000)
```


```{r alpha-vs-beta-diversity, fig.width=12, fig.height=5}
alphadiver_cohort2_subj <- alphadiver_cohort2 %>%
  group_by(SubjectID) %>%
  dplyr::summarise(mean_alpha_diversity = mean(pdrare))

df_alpha_beta <- beta_diver_df_bl_subj %>%
  left_join(alphadiver_cohort2_subj) 

ggplot(df_alpha_beta, 
       aes(x = mean_alpha_diversity, y = dist)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor(method = "spearman") + 
  facet_wrap(~metric, scales = "free", ncol = 6) + 
  labs(x = "Mean alpha diversity", y = "Pre-diet daily plasticity")
```


```{r alpha-vs-beta-diversity-diet, fig.width=12, fig.height=7}
ggplot(df_alpha_beta, 
       aes(x = mean_alpha_diversity, y = dist)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm") +
  ggpubr::stat_cor(method = "spearman") + 
  facet_wrap(~ Diet +  metric, scales = "free", ncol = 5) + 
  labs(x = "Mean alpha diversity", y = "Pre-diet daily plasticity")
```


# Microbiota plasticity compared to adherence/dietary fidelity

## Adherence/dietary fidelity v weight loss
```{r adherence_plots, fig.height=4, fig.width = 8}
ggplot(beta_diver_df_bySubj, 
       aes(x = Percent.12Mo, y = adherence.changeFromBL, color = Gender)
) + 
  geom_point() + 
  stat_smooth(method = "lm") + 
  facet_wrap(~Diet) + 
  ggpubr::stat_cor(method = "pearson") + 
  labs(y = "Dietary compliance (change from BL)", x = "12-month weight loss (%)")

```


##Microbiota plasticity v. adherence/dietary fidelity 

Pre-diet daily plasticity v adherence/dietary fidelity 


```{r adherence-plasticity-MainFig, fig.height=6, fig.width=8}
gender_cols <- c("Female" = "#74D055FF", 
                  "Male" = brewer.pal(8, "Dark2")[3])
ggplot(beta_diver_df_bySubj %>% 
         filter(VarCat %in% c("BL v. BL", "BL v. 10wk"),
           metric %in% c("Bray-Curtis")),
       aes(y = dist, x = adherence.changeFromBL, 
           group = Gender, color= Gender, fill = Gender, 
           lty = Gender)) +
  geom_smooth(method = "lm", lwd = 1.5, alpha = 0.25) +
  geom_point(size = 3, color = "grey50", pch = 21) + 
  ggpubr::stat_cor(method = "spearman") + 
  facet_grid(VarCat ~ Diet, scale ="free") +
  labs(x = "Dietary compliance", 
       y = "Microbiota plasticity", 
       color = "Gender") +
  scale_linetype_manual(values = c(2, 1)) +
  scale_color_manual(values = gender_cols) + 
  scale_fill_manual(values = gender_cols) + 
  theme(legend.position = "top", text = element_text(size = 20))
```

```{r adherence-plasticity-SexTogether, fig.height=6, fig.width=8}
#Not separated by gender
ggplot(beta_diver_df_bySubj %>% 
         filter(VarCat %in% c("BL v. BL", "BL v. 10wk"),
           metric %in% c("Bray-Curtis")),
       aes(y = dist, x = adherence.changeFromBL)) +
  geom_smooth(method = "lm", lwd = 1.5, alpha = 0.25) +
  geom_point(size = 3, color = "grey50", pch = 21) + 
  ggpubr::stat_cor(method = "spearman") + 
  facet_grid(VarCat ~ Diet, scale ="free") +
  labs(x = "Dietary compliance", 
       y = "Microbiota plasticity") +
  scale_linetype_manual(values = c(2, 1)) +
  # scale_color_manual(values = gender_cols) + 
  # scale_fill_manual(values = gender_cols) + 
  theme(legend.position = "top", text = element_text(size = 20))

```




```{r adherence-plasticity-bl-SI, fig.height=12, fig.width=6}
ggplot(beta_diver_df_bySubj %>% 
         filter(VarCat == "BL v. BL"),
       aes(y = dist, x = adherence.changeFromBL, group = Gender, 
           color= Gender, fill = Gender, lty = Gender)) +
  geom_smooth(method = "lm", lwd = 1.5, alpha = 0.25) +
  geom_point(size = 3, color = "grey50", pch = 21) + 
  ggpubr::stat_cor(method = "spearman") + 
  facet_grid(metric ~ Diet, scale = "free") +
  labs(x = "Dietary compliance", 
       y = "Pre-diet microbiota plasticity", 
       color = "Gender") +
  scale_linetype_manual(values = c(2, 1)) +
  scale_color_manual(values = gender_cols) + 
  scale_fill_manual(values = gender_cols) + 
  theme(legend.position = "top", text = element_text(size = 20))
```



### Pre-diet to 10-week plasticity v adherence/dietary fidelity 

```{r adherence-plasticity-bl_10wk_SI, fig.height=12, fig.width=6}
ggplot(beta_diver_df_bySubj %>% 
         filter(
           VarCat == "BL v. 10wk"),
       aes(y = dist, x = adherence.changeFromBL, group = Gender, 
           color= Gender, fill = Gender, lty = Gender)) +
  geom_smooth(method = "lm", lwd = 1.5, alpha = 0.25) +
  geom_point(size = 3, color = "grey50", pch = 21) + 
  ggpubr::stat_cor(method = "spearman") + 
  facet_grid(metric ~ Diet, scale = "free") +
  labs(x = "Dietary compliance", 
       y = "Microbiota plasticity between pre-diet and 10-weeks", 
       color = "Gender") +
  scale_linetype_manual(values = c(2, 1)) +
  scale_color_manual(values = gender_cols) + 
  scale_fill_manual(values = gender_cols) + 
  theme(legend.position = "top", text = element_text(size = 20))
```




## Calculating variability on tip-agglomerated dataset

## Load data

```{r}
(pstips <- readRDS("../data/phyloseq/diet_participants_tipglom_with_adherence.rds"))

# calculate adherence, and Subj_Group variables
newSam_data <- sample_data(pstips) %>%
  data.frame() %>%
  mutate_at(vars(starts_with("adherence")), as.numeric) %>%
  mutate(adherence.raw = rowMeans(select(., adherence_3_months, adherence_6_months, adherence_12_months), na.rm = T),
         adherence.pct = 100-adherence.raw, 
         adherence.changeFromBL = adherence_baseline - adherence.raw,
         Subj_Group = factor(paste0(SubjectID, "_", Timepoint)), 
         rowname = MeasID, 
         Cat.12Mo = factor(Cat.12Mo, levels = c("US", "MS", "VS"))) %>%
  column_to_rownames("rowname")

sample_data(pstips) <- newSam_data


minSampleDepth <- 10000
cohort         <- 2

(pstips <- subset_samples(pstips, sample_sums(pstips) > minSampleDepth))
(pstips <- subset_samples(pstips, Cohort == cohort))
(pstips <- subset_samples(pstips, !(SubjectID == "1139" & Timepoint == "10 Week")))
(pstips <- subset_samples(pstips, !is.na(Cat.12Mo)))

(pstips <- subset_taxa(pstips, taxa_sums(pstips) > 0))

```

### Biological samples that were sequenced twice

```{r}
(sample_data(pstips) %>% 
   group_by(SampleID) %>% 
   summarise(
     count = n(), 
     meas = paste0(MeasID, collapse = "; ")
     ) %>%
   filter(count > 1))
```

Combine repeated samples by computing the mean 

```{r}
pstips <- collapse_samples(pstips)
pstips
```


```{r, eval = FALSE}

dist_lst <- c("jaccard", "unifrac", "jsd", "bray", "wunifrac")
beta_diver_lst <- lapply(dist_lst, function(dist) 
  beta_diversity(
    physeq  = pstips,
    metric = dist, 
    timediff_thresh = 500,
    asinh = TRUE))
names(beta_diver_lst) <- dist_lst
saveRDS(beta_diver_lst, file = "../results/beta_diver_tip_glom.rds")
```

```{r}
beta_diver_lst <- readRDS("../results/beta_diver_tip_glom.rds")
beta_diver_df <- plyr::ldply(beta_diver_lst, function(df) df, .id = "metric") %>%
  mutate(metric = factor(metric, levels = dist_lst, labels = c("Jaccard", "UniFrac", "Jensen-Shannon", "Bray-Curtis", "Weighted UniFrac")),
         Cat.12Mo = factor(Cat.12Mo, levels = c("US", "MS", "VS")))
```

##Daily baseline (pre-diet) variability
Keeping only consecutive samples that are within 5 days of each other 


```{r}
beta_diver_df_bl <- beta_diver_df %>%
  filter(Timepoint_1 == Timepoint_2,
         Timepoint_1 == "Baseline",
         abs(TimeDiff) <= 5) %>%
  group_by(SubjectID, metric) %>%
  top_n(2, wt = -TimeDiff) 


beta_diver_df_bl_subj <- beta_diver_df_bl %>%
  data.frame() %>%
  group_by(metric, SubjectID, Diet, Gender) %>%
  dplyr::summarise(
    dist = mean(dist),
    Cat.12Mo = unique(Cat.12Mo), Percent.12Mo = unique(Percent.12Mo))


```

```{r short-term-var-diet-tips, fig.height=12, fig.width=6}
ggplot(beta_diver_df_bl_subj) +
  geom_boxplot(
    aes(y = dist, x = Cat.12Mo, color = Diet)
  ) +
  geom_jitter(
    data = beta_diver_df_bl,
    aes(y = dist, x = Cat.12Mo),
    width = 0.1, 
    height = 0,
    color = "grey55", 
    size = 1
  ) +
  geom_point(
    aes(y = dist, x = Cat.12Mo, color = Diet),
    shape = 17, 
    size = 3
  ) +
  stat_signif(
    data = beta_diver_df_bl_subj, 
    aes(y = dist, x = Cat.12Mo),
    stat = "signif",
    comparisons = list(c("VS", "US")),
    test = "wilcox.test", 
#    test.args = list(alternative = "greater"),
    manual = FALSE,
    margin_top = -.12 , 
    map_signif_level = F, 
    vjust = -0.1
  ) +
  facet_grid(metric~Diet, scales = "free") +
  scale_color_manual(values =  diet_cols) + 
  labs(y = "Pre-diet daily plasticity", 
       x = "12-month weight-loss success category") +
  theme(legend.position="none")
```

## Microbiota plasticity from pre-diet to 10 weeks
```{r}
beta_diver_df_longTerm <- beta_diver_df %>%
  filter(Timepoint_1 != Timepoint_2) 


beta_diver_df_longTerm_subj <- beta_diver_df_longTerm %>%
  group_by(metric, SubjectID, Diet, Cat.12Mo,Percent.12Mo) %>%
  dplyr::summarise(dist = mean(dist))
```


```{r}
table(beta_diver_df_longTerm_subj$Diet, beta_diver_df_longTerm_subj$metric)
```


```{r long-term-var-diet-tips_pairwise, fig.height=12, fig.width=6}
ggplot(data = beta_diver_df_longTerm_subj) +
  geom_boxplot(
    aes(y = dist, x = Cat.12Mo, color = Diet)
  ) +
  geom_jitter(
    data = beta_diver_df_longTerm,
    width = 0.1, height = 0,
    aes(y = dist, x = Cat.12Mo),
    color = "grey55", size = 1
  ) +
  geom_beeswarm(
    aes(y = dist, x = Cat.12Mo, color = Diet),
    shape = 17, 
    size = 3
  ) +
  stat_signif(
    aes(y = dist, x = Cat.12Mo),
    stat = "signif", 
    comparisons = list(c("VS", "US")), 
    test = "wilcox.test", 
#    test.args = list(alternative = "greater"), 
    manual = FALSE, 
    margin_top = -.12 , 
    map_signif_level = FALSE, 
    vjust = -0.1
  ) + 
  facet_grid(metric~Diet, scales = "free") +
  scale_color_manual(values =  diet_cols) + 
  labs(y = "Microbiota plasticity between pre-diet and 10-weeks",
       x = "12-month weight loss success category") + 
  theme(legend.position="none")  
  
```
 

## Comparison of plasticity across all time points

```{r}
beta_diver_df <- beta_diver_df %>%
  mutate(
    VarCat = factor(
      ifelse(Timepoint_1 == Timepoint_2, 
             ifelse(Timepoint_1 == "Baseline", "BL v. BL",
                    "10wk v. 10wk"), "BL v. 10wk"),
      levels = c("BL v. BL", "10wk v. 10wk", "BL v. 10wk"))) %>%
  filter(!(Timepoint_1 == Timepoint_2 & TimeDiff >5))

beta_diver_df_bySubj <- beta_diver_df %>%
  group_by(SubjectID, Gender, Diet, VarCat, metric, Cat.12Mo, Percent.12Mo, adherence.pct, adherence.changeFromBL) %>%
  dplyr::summarise(dist = mean(dist), meanTimeDiff = mean(TimeDiff), medianTimeDiff = median(TimeDiff))

(beta_diver_Summary <- beta_diver_df_bySubj %>%
  filter(metric == "Bray-Curtis") %>%
  group_by(Diet, VarCat) %>%
  dplyr::summarise(meanTD = mean(meanTimeDiff), medianTD = median(meanTimeDiff)))

```


```{r variability_acrossAllTimes-tip_glom, fig.height=5, fig.width=12}
ggplot(
  data = beta_diver_df_bySubj %>% filter(metric %in% c("Bray-Curtis")), 
  aes(x = VarCat, y = dist)
) + 
  geom_boxplot(
    aes(color = Cat.12Mo),
    outlier.shape = NA
  ) + 
  geom_point(
    data = beta_diver_df %>% filter(metric %in% c("Bray-Curtis")), 
    aes( x = VarCat, y = dist, fill = Cat.12Mo), 
    position = position_jitterdodge(jitter.width = 0.25), 
    color = "grey55", size = 1.5, alpha = 0.5
  ) + 
  geom_point(
    aes(color = Cat.12Mo),
    position = position_jitterdodge(jitter.width = 0.1), 
    shape = 17, size = 3
  ) +
  stat_compare_means(
    data = beta_diver_df_bySubj %>% 
      filter(metric %in% c("Bray-Curtis"), Cat.12Mo != "MS"),
    mapping = aes(group = Cat.12Mo),
    label.y.npc = 0.9) +
  facet_wrap(~Diet, scales = "free", ncol = 2) + 
  scale_color_manual(values = c("#E41A1C", "#377EB8", "#4DAF4A")) + 
  labs(x = " ", y = "Bray-Curtis dissimilarity", color = "") + 
  coord_flip() + 
  guides(colour = guide_legend(reverse=T, title = "Weight Loss Success")) +
  theme(text = element_text(size = 20), legend.position = "top")

```



# Microbiota plasticity compared to adherence/dietary fidelity

##Microbiota plasticity v. adherence/dietary fidelity 

```{r adherence-plasticity-bl-tip_glom, fig.height = 8, fig.width=7}
ggplot(beta_diver_df_bySubj %>% 
         filter(VarCat %in% c("BL v. BL", "BL v. 10wk"),
           metric %in% c("Bray-Curtis")),
       aes(x = dist, y = adherence.changeFromBL, group = Gender, 
           color = Gender, fill = Gender, lty = Gender)) +
  geom_smooth(method = "lm", lwd = 1.5, alpha = 0.25) +
  geom_point(size = 3, color = "grey50", pch = 21) + 
  ggpubr::stat_cor(aes(color = Gender), method = "spearman") + 
  facet_wrap(VarCat ~ Diet) +
  labs(y = "Dietary compliance", 
       x = "Microbiota plasticity (Bray-Curtis dissimilarity)") +
  scale_linetype_manual(values = c(2, 1)) +
  scale_color_manual(values = gender_cols) + 
  scale_fill_manual(values = gender_cols) + 
  theme(legend.position = "top", text = element_text(size = 20))
```




```{r adherence-plasticity-bl-tip_glom_SI,fig.height=12, fig.width=6}
ggplot(
  beta_diver_df_bySubj %>% filter(VarCat == "BL v. BL"),
  aes(y = dist, x = adherence.changeFromBL, group = Gender, 
      color= Gender, fill = Gender, lty = Gender)) +  
  geom_smooth(method = "lm", lwd = 1.5, alpha = 0.25) +
  geom_point(size = 3, color = "grey50", pch = 21) + 
  ggpubr::stat_cor(method = "spearman") + 
  facet_grid(metric ~ Diet, scales = "free") +  
  labs(x = "Dietary compliance", 
       y = "Pre-diet microbiota plasticity", 
       color = "Gender") +
  scale_linetype_manual(values = c(2, 1)) +
  scale_color_manual(values = gender_cols) + 
  scale_fill_manual(values = gender_cols) + 
  theme(legend.position = "top", text = element_text(size = 20))
```

```{r  adherence-plasticity-bl-tip_glom_SI_long, fig.height=12, fig.width=6}
ggplot(beta_diver_df_bySubj %>% 
         filter(
           VarCat == "BL v. 10wk"),
       aes(y = dist, x = adherence.changeFromBL, 
           group = Gender, color= Gender, fill = Gender, 
           lty = Gender)) +
  geom_smooth(method = "lm", lwd = 1.5, alpha = 0.25) +
  geom_point(size = 3, color = "grey50", pch = 21) + 
  ggpubr::stat_cor(method = "spearman") + 
  facet_grid(metric ~ Diet, scales = "free") +
  scale_linetype_manual(values = c(2, 1)) +
  scale_color_manual(values = gender_cols) + 
  scale_fill_manual(values = gender_cols) + 
  theme(legend.position = "top", text = element_text(size = 20))  +
  labs(x = "Dietary compliance",
       y = "Microbiota plasticity between pre-diet and 10-weeks", 
       color = "Gender")
```


```{r}
sessionInfo()
```